Short and Mechanized Logical Relation for Dependent Types

### 2024-02-26 paper/paper.tex
Shrink to 25 pages

%%
%% The "title" command has an optional parameter,
%% allowing the author to define a "short title" to be used in page headers.
\title{Short and Mechanized Logical Relation for Dependent Type Theory}

@@

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
Proof by logical relation is a powerful technique that has been used
to derive metatheoretic properties of type systems, such as
consistency and parametricity. While there exists a
plethora of introductory materials about logical relation in the
context of simply typed or polymorphic lambda calculus, a streamlined
presentation of proof by logical relation for a dependently language
is lacking. In this paper, I present a short
consistency proof for a dependently typed language that contains a
rich set of features, including a full cumulative universe
hierarchy, booleans, and an intensional identity type. We have
fully mechanized the consistency proof using the Coq proof assistant
in under 1000 lines of code.
\end{abstract}

@@

Our proof, designed to capture the essence of this proof technique in this
setting, requires less than 1000 lines of code. We have achieved this
significant reduction through a number of means: the careful selection of the
features that we include in the object type type system and the results that
we prove about it, in addition to the judicious use of automation.  Our
language is small, but includes enough to be illustrative. For example, we
eschew inductive datatypes or W-types, the but we do include propositionality
equality and booleans in order to capture the challenges presented by indexed
types and dependent pattern matching. We do not show the decidability of type
checking, nor do we develop a PER semantics, but we do demonstrate how our
consistency proof can be extended (at a moderate cost of 500 lines of code) to
show the existence of $\beta\eta$-normal form for well-typed open \emph{and}
closed terms. We include a full predicative universe hiercharchy and
type-level computation, to demonstrate the logical strength of the
approach. % \scw{Say that we don't have impredicativity, but neither does anyone
  % else?}\yl{Nobody else but Barras. I'll find the citation}
%%
%% SCW: save specific comparisons with other systems for later
%%
% The big difference in the size of the developments does not
% necessarily imply that our proof technique leads to a more concise
% proof due to the differing expressiveness of the languages and the
% different metatheoretic results being established. For
% example, \citet{anand2014towards} mechanizes the metatheory of
% Nuprl~\citep{constable1986implementing}, which, on top of all the
% features that \lang{} supports, includes W-types and partial types and
% requires a PER semantics to model its extensional typed
% equality. The object language from \citet{decagda} does not support
% identity types and only has one predicative universe, but additionally
% supports $\Sigma$ types. However, for a researcher who wishes to learn
% the underlying techniques for mechanized logical relation for
% dependent type theory, a small development like ours is much easier to
% navigate and understand. Furthermore, we show how our simple
% consistency proof can be easily extended to show the existence of
% $\beta\eta$-normal form for well-typed open \emph{and} closed terms,
% giving us a metatheoretic result almost as strong as the one from
% \citet{decagda} at the moderate cost of around 500 lines of extra code.

@@

In Section~\ref{sec:spec}, we introduce \lang{}, the object language
of interest, and establish some elementary properties about parallel
reduction, which will be used as the equational theory of our language
and play a central role in defining our logical relation. The design
of \lang{} is reminiscent of Martin-Löf style predicatvie type
systems. The untyped conversion rule of \lang{} is inspired by Pure
Type Systems~\citet{barendregt1991introduction}, though we discuss how
our proof can be easily adjusted for systems with typed judgmental
equality.  We design \lang{} to incorporate some of the most common
features of dependent type theory, including an intensional identity
type, a boolean type, large elimination, dependent elimination, and a
full universe hierarchy. We believe a good coverage of features will
make our proof applicable to a broad range of type systems.

@@

\begin{figure}[h]
\begin{minipage}{0.9\textwidth}
\drules[P]{$[[a => b]]$}{Parallel Reduction}{AppAbs, IfTrue, IfFalse, JRefl}
\drules[PS]{$[[a =>+ b]]$}{Transitive Closure of Parallel Reduction}{Refl, Step}
\drules[C]{$[[a <=> b]]$}{Convertibility}{Intro}
\end{minipage}
\caption{Parallel reduction ($\beta$-rules only) and convertibility }
\label{fig:par}
\end{figure}

@@

In \rref{T-Var}, $[[G n]]$ is the partial function defined
by the equations $[[ (G ++ A)  0 ]] := [[A]]$ and $[[ (G ++ A) Suc
n ]] := [[G n]]$.
The precondition $[[n < | G | ]]$, where $[[ | G |
]]$ represents the length of the context $[[G]]$, ensures that $[[G
n]]$ has a defined output. The type $[[G n]]$ needs to be shifted
by $[[Suc n]]$ since the types stored in the context $[[G]]$ are
scoped differently. In the typing judgment $[[G |- a : A]]$, we want
both $[[a]]$ and $[[A]]$ to have the same scope. However, the closer a
type is to the head of $[[G]]$, the smaller de Bruijn it has since
there is less variables in the context it can refer to (see \rref{Ctx-Cons}). Shifting $[[G
n]]$ by $[[Suc n]]$ weakens the type so it is scoped consistently with
the full context $[[G]]$ rather than a truncated version of
$[[G]]$. In \rref{T-J}, the elimination form for the idnetity type,
$[[a]]$ and $[[A]]$ need to be shifted in the context for a similar
reason; an extra type $[[A]]$ appears after the $[[G]]$ and therefore
the terms $[[a]]$ and $[[A]]$ in the identity type both need to be shifted
to refer to the same binders.

@@

The complexity of applying proof by logical relation to dependent types stem
from the fact that the logical relation is much harder to define. In
simply typed languages, the logical relation is defined as a recursive
function over the type $[[A]]$.
In dependent types, the type
$[[A]]$ can take the form $[[(\ var 0) Bool]]$. To assign meaning to
this type, we need to first reduce it to $[[Bool]]$. However, we
cannot write a function that performs the reduction because we do not
know the termination of well-typed terms a priori. As a result, we
define the logical relation as an inductively defined relation,
reminiscent of how we specify the reduction graph of a partial
function; the functionality of the relation can later be recovered in Lemma~\ref{lemma:logreldeter}.

@@

We say that a set of terms $[[S]]$ is closed under expansion if given
$[[a in S]]$, then $[[b in S]]$ for all $[[b => a]]$.
The final property we want to show is that the output set $[[S]]$ from
the logical relation is closed under expansion. Unlike the previous
lemmas, we directly state the lemma
in terms of $[[InterpR i A S]]$ rather than $[[Interp I i A S]]$
because we need to know something about $[[I]]$ for this property to
hold in the \rref{I-Set} case.

\yl{Can lift this lemma somewhere earlier in the section since it doesn't really
  depend on anything}
\begin{lemma}[Interpreted sets are closed under expansion\footnote{\dotv{semtyping.v}{InterpExt\_back\_clos}}]
  \label{lemma:logrelbackclos}
  If $[[InterpR i A S]]$, then the set $[[S]]$ is closed under expansion.
\end{lemma}
\begin{proof}
  By the definition of $[[InterpR i A S]]$ from
  Figure~\ref{fig:logrelrec}, we unfold $[[InterpR i A S]]$ by one
  step into $[[ Interp I i A S  ]]$ where $[[I i]] := [[{A | exists S
    , InterpR i A S}]]$.
  We then proceed by induction over the derivation of $[[Interp I i A S]]$.

@@

\begin{definition}[Semantic well-formed substitution\footnote{\dotv{soundness.v}{$\rho$\_ok}}]
Define $[[rho |= G]]$ when
\[ \forall i\ j\ S, \text{ if }[[i < |G|]]\text{ and
                     } [[InterpR j (G i < up Suc i > ) { rho } S ]] \text{, then } [[rho i in S]] \]
\end{definition}

@@

The $[[rho |= G]]$ notation denotes the semantic well-formedness of a
substitution $[[rho]]$ with respect to a context $[[G]]$. Intuitively, for any
variable $[[i]]$ in the context (thus the restriction $[[i < | G |]]$),
$[[rho i]]$ is a term that inhabits all possible interpretations of the type
$[[G i]]$ from the context.

Since we are in a dependently typed system, there
may be variables in $[[G i]]$, too. The type $[[(G i < up Suc i > ) {
  rho }]]$ applies $[[rho]]$ to replace those variables. The shifting
operation $[[up Suc i]]$ can be viewed as an artifact of the de Bruijn
representation to lift the type $[[G i]]$ to the right scope before
substitution can be carried out.

Finally, despite the for all quantification over the interpreted set
$[[S]]$ and the universe level $[[j]]$ in the definition of $[[rho |=
G]]$, we only need show that there exists some $[[j]]$ and $[[S]]$
such that $[[InterpR j (G i < up Suc i > ) { rho } S]]$ and $[[rho i
in S]]$. This idea is encapsulated in the second of the two structural
law about $[[rho |= G]]$.
\begin{lemma}[Well-formed $[[rho]]$ empty\footnote{\dotv{soundness.v}{$\rho$\_ok\_nil}}]
  \label{lemma:rhowfempty}
  $[[rho |= G]]$ whenever $[[G]]$ is the empty context.
\end{lemma}
\begin{lemma}[Well-formed $[[rho]]$ cons\footnote{\dotv{soundness.v}{$\rho$\_ok\_cons}}]
  If $[[InterpR i A S]]$, $[[a in S]]$, and $[[rho |= G]]$, then
  $[[rho .: a |= G ++ A]]$.
\end{lemma}

@@

\begin{definition}[Semantic context well-formedness\footnote{\dotv{soundness.v}{SemWff}}]
Define $[[|= G]]$ when
\[
   \forall [[i < |G|]],\text{ there exists some } [[j]] \text{ such that } [[drop Suc i G |= G i : Set j]] \]
\end{definition}

In this definition, we use the notation $[[drop i G]]$ to denote
the typing context obtained by dropping the last $[[i]]$ elements of
$[[G]]$. When $[[G]]$ has fewer than $[[i]]$ elements, $[[drop i G]]$ returns
the empty list. Intuitively, $[[|= G]]$ says that all
types in the context must correspond to some set according to the logical
relation. The truncation to the context $[[G]]$ is due to the telescopic
scoping of the types in the context. An alternative approach would be to define $[[|= G]]$
inductively, similar to $[[|- G]]$, but we find it easier to use this definition and
recover the structural rules as lemmas.
\begin{lemma}[Semantic context well-formedness cons]
  \label{lemma:semwffcons}
  If $[[|= G]]$ and $[[G |= A : Set i]]$, then $[[|= G ++ A]]$.
\end{lemma}
\begin{lemma}[Semantic context well-formedness empty]
  \label{lemma:semwffempty}
  $[[|= G]]$ holds when $[[G]]$ is empty.
\end{lemma}

@@

Next, we show some non-trivial cases of the fundamental theorem as
top-level lemmas. For example, we can define the semantic analogue to the
syntactic typing rule for variables (\rref{T-Var}).
\begin{lemma}[ST-Var]
  \label{lemma:stvar}
  If $[[|= G]]$ and $[[i < |G|]]$, then $[[G |= var i : G i < up Suc i  >]]$.
\end{lemma}
\begin{proof}
  Suppose $[[rho |= G]]$. By the definition of semantic typing, we
need to show that there exists some $[[j]]$ and $[[S]]$ such that
  \begin{itemize}
  \item $[[InterpR j G i < up Suc i > { rho } S]]$
  \item $[[rho i in S]]$
  \end{itemize}
  From the definition of $[[|= G]]$, we know that there exists some
  $[[j]]$ such that $[[drop Suc i G |= G i : Set j]]$. This is not
  immediately useful since $[[rho |= G]]$, not the truncated
  context. We claim, without giving the proof, that from $[[drop Suc i
  G |= G i  : Set j]]$, we can derive $[[G |= G i {up Suc i} : Set
  j]]$. A rigorous justification would require us to formally state
  and show renaming/weakening lemma for semantic typing. We instead
  provide an informal justification since the proof of renaming
  is uninteresting and only requires unfolding definitions, but its
  statement introduces unnecessary line noise related to shifting.

  Intuitively, a substitution $[[rho]]$
  that closes over a context should always induce a truncated
  substituion for each truncated context. As a result, if
  some term $[[b]]$ is semantically well-typed under some truncated
  context $[[G]]$, then it must also be semantically well-typed under
  the context $[[G]]$ up to shifting.
%
  Knowing that $[[G |= G i {up Suc i} : Set j]]$ holds, we can then
  apply Lemma~\ref{lemma:setinv} to show that
  $[[InterpR j G i < up Suc i > { rho } S]]$. It now
  suffices to show $[[rho i in S]]$, but that is immediate from the
  definition of $[[rho |= G]]$.
\end{proof}

@@

\begin{lemma}[ST-Set]
  \label{lemma:stset}
  If $[[i < j]]$, then $[[G |= Set i : Set j]]$.
\end{lemma}
\begin{proof}
  Immediate by Lemma~\ref{lemma:setinv} and \rref{I-Set}.
\end{proof}

\begin{lemma}[ST-Pi]
  \label{lemma:stpi}
  If $[[G |= A : Set i]]$ and $[[G ++ A |= B : Set i]]$, then $[[G |= Pi
  A B : Set i]]$.
\end{lemma}
\begin{proof}
  Applying Lemma~\ref{lemma:setinv} to the
  conclusion, it now suffices to show that given $[[rho |= G]]$, there
  exists some $[[S]]$ such that $[[InterpR i Pi A{rho} B{up rho} S]]$.
  From Lemma~\ref{lemma:setinv} and $[[G |= A : Set i]]$, we know that
  there exists some set $[[S0]]$ such that $[[InterpR i A {rho} S0]]$.
From $[[G ++ A |= B : Set i]]$, we know that there must
exists $[[S]]$ such that $[[InterpR i B {rho .: a} S]]$ for every $[[a
in S0]]$. The conclusion immediately follows from Lemma~\ref{lemma:piintroalt}.
\end{proof}

@@

  By Lemma~\ref{lemma:setinv} and the premise $[[G |= Pi A B : Set
  i]]$, there exists some set $[[S]]$ such that
  $[[InterpR i Pi A {rho} B {up rho} S]]$. It now suffices to show that
  $[[\A{rho} b{up rho} in S
  ]]$. By Lemma~\ref{lemma:piinvalt}, there exists some $[[S0]]$ such
  that all following conditions hold:
  \begin{itemize}
  \item $[[InterpR i A{rho} S0]]$
  \item $[[forall a, (# a in S0 implies (# exists S1 , InterpR i B
    {rho .: a}
    S1 #) #)]]$
  \item $[[S = { b | forall a, (# a in S0 , forall
      S1, (# InterpR i B {rho .: a} S1,  b a in S1 #) #) }]]$
  \end{itemize}
  To show that $[[\A{rho} b{up rho} in S]]$, we need to prove
  that given $[[a in S0]]$,
  $[[Interp I i B {rho .: a} S1]]$, we have  $[[( \A{rho} b{up rho} )
  a in S1]]$.
  By Lemma~\ref{lemma:logrelbackclos}, the set $[[S1]]$ is closed
  under expansion. Since $[[( \A{rho} b{up rho} )
  a => b {up rho} {a}]] = [[b {rho .: a}]]$, it suffices to show that
  $[[b {rho .: a} in S1]]$, which is immediate from $[[G ++ A |= b :
  B]]$ and the fact that the logical relation is deterministic and
  cumulative (Lemma~\ref{lemma:logreldeterhet}).
\end{proof}

@@

\begin{theorem}[The Fundamental Theorem\footnote{\dotv{soundness.v}{soundness}}]
  \label{theorem:soundness}\leavevmode
  \begin{itemize}
  \item If $[[G |- a : A]]$, then $[[G |= a : A]]$.
  \item If $[[|- G]]$, then $[[|= G]]$.
  \end{itemize}
\end{theorem}
\begin{proof}
  Proof by mutual induction over the derivation of $[[G |- a :
  A]]$ and $[[|- G]]$.   The cases related to context well-formedness immediately follows
  from Lemma~\ref{lemma:semwffcons}. The semantic typing rules
(Lemmas~\ref{lemma:stvar},~\ref{lemma:stset},~\ref{lemma:stpi},~\ref{lemma:stabs},~\ref{lemma:stapp})
  can be used to discharge their syntactic counterparts
  (e.g. Lemma~\ref{lemma:stabs} for case \rref{T-Abs}). The remaining
  cases not covered by the lemmas are similar to the ones already
  shown.
\end{proof}

@@

Furthermore, Lemma~\ref{lemma:parnenf}, in its current weaker form, would
still hold after we extend our equational theory with the function $\eta$
rule, where parallel reduction can take $\eta$ steps but still preserves
$\beta\eta$-normal form.

@@

Furthermore, to prove this adequacy property, we need to strengthen it to also
give us more information about neutral terms as we proceed by induction. In
particular, we need to know that all terms that reduce to neutral forms are
contained within the interpretation.
\scw{Also, cite where this form of proof originally
appeared? This part is standard.}
Therefore, we formally define when a set is a \emph{reducibility candidate}
(shortened as $CR$) as follows.
\begin{definition}[Reducibility Candidates (CR)\footnote{\dotv{semtypingopen.v}{CR}}]
  Let $[[S]]$ be a set of lambda terms. We say that $[[S]] \in CR$
  if and only if conditions $CR_1$ and $CR_2$ hold.
  \begin{itemize}
  \item $[[S]] \in CR_1 \iff\ [[forall a, (#  wne a implies a in S #)]]$
  \item $[[S]] \in CR_2 \iff\ [[forall a, (# a in S implies wn a #)]]$
  \end{itemize}
\end{definition}

@@

The automation provided by CoqHammer not only gives us a proof that is shorter
and more resiliant to changes, but also gives useful documentation for readers
who wish to understand how the underlying proof. Although automation performs
extensive search, it does not use lemmas or invert derivations that are not
specified in the \texttt{use} or \texttt{inv} flag.

>>>>

%%
%% The "title" command has an optional parameter,
%% allowing the author to define a "short title" to be used in page headers.
\title{Short and Mechanized Logical Relation for Dependent Types}

@@

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
Proof by logical relation is a powerful technique that has been used
to derive metatheoretic properties of type systems, such as
consistency and parametricity. While there exists a
plethora of introductory materials about logical relation in the
context of simply typed or polymorphic lambda calculus, a streamlined
presentation of proof by logical relation for a dependently language
is lacking. In this paper, we present a short
consistency proof for a dependently typed language that contains a
rich set of features, including a full universe
hierarchy, booleans, and an intensional identity type. We show that
the logical relation can be easily extended to prove the existence of
$\beta\eta$-normal forms.
We have
fully mechanized the consistency proof using the Coq proof assistant
in under 1000 lines of code, with 500 lines of additional code for the
$\beta\eta$-normal form extension.
\end{abstract}

@@

Our proof, designed to capture the essence of this proof technique in this
setting, requires less than 1000 lines of code. We have achieved this
significant reduction through a number of means: the careful selection of the
features that we include in the object type type system and the results that
we prove about it, in addition to the judicious use of automation.  Our
language is small, but includes enough to be illustrative. For example, we
eschew inductive datatypes or W-types, but we do include propositionality
equality and booleans in order to capture the challenges presented by indexed
types and dependent pattern matching. We do not show the decidability of type
checking, nor do we develop a PER semantics, but we do demonstrate how our
consistency proof can be extended (at a moderate cost of 500 lines of code) to
show the existence of $\beta\eta$-normal form for well-typed open \emph{and}
closed terms. We include a full predicative universe hiercharchy and
type-level computation, to demonstrate the logical strength of the
approach. % \scw{Say that we don't have impredicativity, but neither does anyone
  % else?}\yl{Nobody else but Barras. I'll find the citation}
%%
%% SCW: save specific comparisons with other systems for later
%%
% The big difference in the size of the developments does not
% necessarily imply that our proof technique leads to a more concise
% proof due to the differing expressiveness of the languages and the
% different metatheoretic results being established. For
% example, \citet{anand2014towards} mechanizes the metatheory of
% Nuprl~\citep{constable1986implementing}, which, on top of all the
% features that \lang{} supports, includes W-types and partial types and
% requires a PER semantics to model its extensional typed
% equality. The object language from \citet{decagda} does not support
% identity types and only has one predicative universe, but additionally
% supports $\Sigma$ types. However, for a researcher who wishes to learn
% the underlying techniques for mechanized logical relation for
% dependent type theory, a small development like ours is much easier to
% navigate and understand. Furthermore, we show how our simple
% consistency proof can be easily extended to show the existence of
% $\beta\eta$-normal form for well-typed open \emph{and} closed terms,
% giving us a metatheoretic result almost as strong as the one from
% \citet{decagda} at the moderate cost of around 500 lines of extra code.

@@

In Section~\ref{sec:spec}, we introduce \lang{}, the object language
of interest, and establish some elementary properties about parallel
reduction, which will be used as the equational theory of our language
and play a central role in defining our logical relation. The design
of \lang{} is reminiscent of Martin-Löf style predicative type
systems. The untyped conversion rule of \lang{} is inspired by Pure
Type Systems~\citet{barendregt1991introduction}, though we discuss how
our proof can be easily adjusted for systems with typed judgmental
equality.  We design \lang{} to incorporate some of the most common
features of dependent type theory, including an intensional identity
type, a boolean type, large elimination, dependent elimination, and a
full universe hierarchy. We believe a good coverage of features will
make our proof applicable to a broad range of type systems.

@@

\begin{figure}[h]
\begin{minipage}{0.9\textwidth}
\drules[P]{$[[a => b]]$}{Parallel Reduction}{AppAbs, IfTrue, IfFalse, JRefl}
% \drules[PS]{$[[a =>+ b]]$}{Transitive Closure of Parallel Reduction}{Refl, Step}
% \drules[C]{$[[a <=> b]]$}{Convertibility}{Intro}
\end{minipage}
\caption{Parallel reduction ($\beta$-rules only)  }
\label{fig:par}
\end{figure}

@@

In \rref{T-Var}, $[[G n]]$ is the partial function defined
by the equations $[[ (G ++ A)  0 ]] := [[A]]$ and $[[ (G ++ A) Suc
n ]] := [[G n]]$.
The precondition $[[n < | G | ]]$, where $[[ | G |
]]$ represents the length of the context $[[G]]$, ensures that $[[G
n]]$ has a defined output. The type $[[G n]]$ needs to be shifted
by $[[Suc n]]$ since the types stored in the context $[[G]]$ are
scoped differently. In the typing judgment $[[G |- a : A]]$, we want
both $[[a]]$ and $[[A]]$ to have the same scope. However, the closer a
type is to the head of $[[G]]$, the smaller de Bruijn it has since
there is less variables in the context it can refer to (see \rref{Ctx-Cons}). Shifting $[[G
n]]$ by $[[Suc n]]$ weakens the type so it is scoped consistently with
the full context $[[G]]$ rather than a truncated version of
$[[G]]$. In \rref{T-J}, the elimination form for the identity type,
the terms $[[a]]$ and $[[A]]$ need to be shifted in the context for a similar
reason; an extra type $[[A]]$ appears after the $[[G]]$ and therefore
the terms $[[a]]$ and $[[A]]$ in the identity type both need to be shifted
to refer to the same binders.

@@

The complexity of applying proof by logical relation to dependent types stems
from the fact that the logical relation is much harder to define. In
simply typed languages, the logical relation is defined as a recursive
function over the type $[[A]]$.
In dependent types, the type
$[[A]]$ can take the form $[[(\ var 0) Bool]]$. To assign meaning to
this type, we need to first reduce it to $[[Bool]]$. However, we
cannot write a function that performs the reduction because we do not
know the termination of well-typed terms a priori. As a result, we
define the logical relation as an inductively defined relation,
reminiscent of how we specify the reduction graph of a partial
function; the functionality of the relation can later be recovered in Lemma~\ref{lemma:logreldeter}.

@@

\begin{definition}[Sets closed under expansion]
  We say that a set of terms $[[S]]$ is closed under expansion if
  given $[[a in S]]$, then $[[b in S]]$ for all $[[b => a]]$.
\end{definition}
The final property we want to show is that the output set $[[S]]$ from
the logical relation is closed under expansion. Unlike the previous
lemmas, we directly state the lemma
in terms of $[[InterpR i A S]]$ rather than $[[Interp I i A S]]$
because we need to know something about $[[I]]$ for this property to
hold in the \rref{I-Set} case.

\yl{Can lift this lemma somewhere earlier in the section since it doesn't really
  depend on anything}
\begin{lemma}[Interpreted sets are closed under expansion\footnote{\dotv{semtyping.v}{InterpUnivN\_back\_clos}}]
  \label{lemma:logrelbackclos}
  If $[[InterpR i A S]]$, then the set $[[S]]$ is closed under expansion.
\end{lemma}
\begin{proof}
  By the definition of $[[InterpR i A S]]$ from
  Figure~\ref{fig:logrelrec}, we unfold $[[InterpR i A S]]$ by one
  step into $[[ Interp I i A S  ]]$ where $[[I i]] := [[{A | exists S
    , InterpR i A S}]]$.
  We then proceed by induction over the derivation of $[[Interp I i A S]]$.

@@

\begin{definition}[Semantic well-formed substitution\footnote{\dotv{soundness.v}{$\rho$\_ok}}]
Define $[[rho |= G]]$ when
\[ \forall n\ i\ S, \text{ if }[[n < |G|]]\text{ and
                     } [[InterpR i (G n < up Suc n > ) { rho } S ]] \text{, then } [[rho n in S]] \]
\end{definition}

@@

The $[[rho |= G]]$ notation denotes the semantic well-formedness of a
substitution $[[rho]]$ with respect to a context $[[G]]$. Intuitively, for any
variable $[[n]]$ in the context (thus the restriction $[[n < | G |]]$),
$[[rho n]]$ is a term that inhabits all possible interpretations of the type
$[[G n]]$ from the context.

Since we are in a dependently typed system, there
may be variables in $[[G n]]$, too. The type $[[(G n < up Suc n > ) {
  rho }]]$ applies $[[rho]]$ to replace those variables. The shifting
operation $[[up Suc n]]$ can be viewed as an artifact of the de Bruijn
representation to lift the type $[[G n]]$ to the right scope before
substitution can be carried out.

Finally, despite the for all quantification over the interpreted set
$[[S]]$ and the universe level $[[i]]$ in the definition of $[[rho |=
G]]$, we only need show that there exists some $[[i]]$ and $[[S]]$
such that $[[InterpR i (G n < up Suc n > ) { rho } S]]$ and $[[rho n
in S]]$ since the logical relation is a partial function (Lemma~\ref{lemma:logreldeter}). This idea is encapsulated in the second of the two structural
law about $[[rho |= G]]$.
\begin{lemma}[Well-formed $[[rho]]$ empty\footnote{\dotv{soundness.v}{$\rho$\_ok\_nil}}]
  \label{lemma:rhowfempty}
  $[[rho |= G]]$ whenever $[[G]]$ is the empty context.
\end{lemma}
\begin{lemma}[Well-formed $[[rho]]$ cons\footnote{\dotv{soundness.v}{$\rho$\_ok\_cons}}]
  If $[[InterpR i A S]]$, $[[a in S]]$, and $[[rho |= G]]$, then
  $[[rho .: a |= G ++ A]]$.
\end{lemma}

@@

\begin{definition}[Semantic context well-formedness\footnote{\dotv{soundness.v}{SemWff}}]
Define $[[|= G]]$ when
\[
   \forall [[n < |G|]],\text{ there exists some } [[i]] \text{ such that } [[drop Suc n G |= G n : Set i]] \]
\end{definition}

In this definition, we use the notation $[[drop n G]]$ to denote
the typing context obtained by dropping the last $[[n]]$ elements of
$[[G]]$. When $[[G]]$ has fewer than $[[i]]$ elements, $[[drop n G]]$ returns
the empty list. Intuitively, $[[|= G]]$ says that all
types in the context must correspond to some set according to the logical
relation. The truncation to the context $[[G]]$ is due to the telescopic
scoping of the types in the context. An alternative approach would be to define $[[|= G]]$
inductively, similar to $[[|- G]]$, but we find it easier to use this definition and
recover the structural rules as lemmas.
\begin{lemma}[Semantic context well-formedness empty]
  \label{lemma:semwffempty}
  $[[|= G]]$ holds when $[[G]]$ is empty.
\end{lemma}
\begin{lemma}[Semantic context well-formedness cons]
  \label{lemma:semwffcons}
  If $[[|= G]]$ and $[[G |= A : Set i]]$, then $[[|= G ++ A]]$.
\end{lemma}

@@

Next, we show some non-trivial cases of the fundamental theorem as
top-level lemmas. For example, we can define the semantic analogue to the
syntactic typing rule for variables (\rref{T-Var}).
\begin{lemma}[ST-Var]
  \label{lemma:stvar}
  If $[[|= G]]$ and $[[n < |G|]]$, then $[[G |= var n : G n < up Suc n  >]]$.
\end{lemma}
\begin{proof}
  Suppose $[[rho |= G]]$. By the definition of semantic typing, we
need to show that there exists some $[[i]]$ and $[[S]]$ such that
  \begin{itemize}
  \item $[[InterpR i G n < up Suc n > { rho } S]]$
  \item $[[rho n in S]]$
  \end{itemize}
  From the definition of $[[|= G]]$, we know that there exists some
  $[[i]]$ such that $[[drop Suc n G |= G n : Set i]]$. This is not
  immediately useful since $[[rho |= G]]$, not the truncated
  context. We claim, without giving the proof, that from $[[drop Suc n
  G |= G n  : Set i]]$, we can derive $[[G |= G n {up Suc n} : Set
  i]]$. A rigorous justification would require us to formally state
  and show renaming/weakening lemma for semantic typing. We instead
  provide an informal justification since the proof of renaming
  is uninteresting and only requires unfolding definitions, but its
  statement introduces unnecessary line noise related to shifting.

  Intuitively, a substitution $[[rho]]$
  that closes over a context should always induce a truncated
  substituion for each truncated context. As a result, if
  some term $[[b]]$ is semantically well-typed under some truncated
  context $[[G]]$, then it must also be semantically well-typed under
  the context $[[G]]$ up to shifting.
%
  Knowing that $[[G |= G n {up Suc n} : Set i]]$ holds, we can then
  apply Lemma~\ref{lemma:setinv} to show that
  $[[InterpR i G n < up Suc n > { rho } S]]$. It now
  suffices to show $[[rho n in S]]$, but that is immediate from the
  definition of $[[rho |= G]]$.
\end{proof}

@@

\begin{lemma}[ST-Set]
  \label{lemma:stset}
  If $[[i < j]]$, then $[[G |= Set i : Set j]]$.
\end{lemma}
\begin{proof}
  Immediate by Lemma~\ref{lemma:setinv} and \rref{IR-Set}.
\end{proof}

\begin{lemma}[ST-Pi]
  \label{lemma:stpi}
  If $[[G |= A : Set i]]$ and $[[G ++ A |= B : Set i]]$, then $[[G |= Pi
  A B : Set i]]$.
\end{lemma}
\begin{proof}
  Applying Lemma~\ref{lemma:setinv} to the
  conclusion, it now suffices to show that given $[[rho |= G]]$, there
  exists some $[[S]]$ such that $[[InterpR i Pi A{rho} B{up rho} S]]$.
  From Lemma~\ref{lemma:setinv} and $[[G |= A : Set i]]$, we know that
  there exists some set $[[S0]]$ such that $[[InterpR i A {rho} S0]]$.
From $[[G ++ A |= B : Set i]]$, we know that there must
exists $[[S]]$ such that $[[InterpR i B {rho .: a} S]]$ for every $[[a
in S0]]$. The conclusion immediately follows from the admissible \rref{I-PiAlt}.
\end{proof}

@@

  By Lemma~\ref{lemma:setinv} and the premise $[[G |= Pi A B : Set
  i]]$, there exists some set $[[S]]$ such that
  $[[InterpR i Pi A {rho} B {up rho} S]]$. It now suffices to show that
  $[[\A{rho} b{up rho} in S
  ]]$. By Lemma~\ref{lemma:piinvalt}, the alternative inversion
  principle for \rref{I-Pi}, there exists some $[[S0]]$ such
  that all following conditions hold:
  \begin{itemize}
  \item $[[InterpR i A{rho} S0]]$
  \item $[[forall a, (# a in S0 implies (# exists S1 , InterpR i B
    {rho .: a}
    S1 #) #)]]$
  \item $[[S = { b | forall a, (# a in S0 , forall
      S1, (# InterpR i B {rho .: a} S1,  b a in S1 #) #) }]]$
  \end{itemize}
  To show that $[[\A{rho} b{up rho} in S]]$, we need to prove
  that given $[[a in S0]]$,
  $[[Interp I i B {rho .: a} S1]]$, we have  $[[( \A{rho} b{up rho} )
  a in S1]]$.
  By Lemma~\ref{lemma:logrelbackclos}, the set $[[S1]]$ is closed
  under expansion. Since $[[( \A{rho} b{up rho} )
  a => b {up rho} {a}]] = [[b {rho .: a}]]$, it suffices to show that
  $[[b {rho .: a} in S1]]$, which is immediate from $[[G ++ A |= b :
  B]]$ and the fact that the logical relation is deterministic and
  cumulative (Lemma~\ref{lemma:logreldeterhet}).
\end{proof}

@@

\begin{theorem}[The Fundamental Theorem\footnote{\dotv{soundness.v}{soundness}}]
  \label{theorem:soundness}\leavevmode
  \begin{itemize}
  \item If $[[G |- a : A]]$, then $[[G |= a : A]]$.
  \item If $[[|- G]]$, then $[[|= G]]$.
  \end{itemize}
\end{theorem}
\begin{proof}
  Proof by mutual induction over the derivation of $[[G |- a :
  A]]$ and $[[|- G]]$.   The cases related to context well-formedness immediately follows
  from Lemmas~\ref{lemma:semwffempty} and \ref{lemma:semwffcons}. The semantic typing rules
(Lemmas~\ref{lemma:stvar},~\ref{lemma:stset},~\ref{lemma:stpi},~\ref{lemma:stabs},~\ref{lemma:stapp})
  can be used to discharge their syntactic counterparts
  (e.g. Lemma~\ref{lemma:stabs} for case \rref{T-Abs}). The remaining
  cases not covered by the lemmas are similar to the ones already
  shown.
\end{proof}

@@

Furthermore, Lemma~\ref{lemma:parnenf}, in its current weaker form, would
still hold after we extend our equational theory with the function $\eta$
rule, where parallel reduction can take $\eta$ steps but still preserves
$\beta$-normal form.

@@

Furthermore, to prove this adequacy property, we need to strengthen it to also
give us more information about neutral terms as we proceed by induction. In
particular, we need to know that all terms that reduce to neutral forms are
contained within the interpretation.
\scw{Also, cite where this form of proof originally
appeared? This part is standard.}
Therefore, we formally define when a set is a \emph{reducibility candidate}
(shortened as $CR$) as follows. Our definition of $CR$ is inspired by
\citet{girard1989proofs}, but not identical since we only care about
weak normalization.
\begin{definition}[Reducibility Candidates (CR)\footnote{\dotv{semtypingopen.v}{CR}}]
  Let $[[S]]$ be a set of lambda terms. We say that $[[S]] \in CR$
  if and only if conditions $CR_1$ and $CR_2$ hold.
  \begin{itemize}
  \item $[[S]] \in CR_1 \iff\ [[forall a, (#  wne a implies a in S #)]]$
  \item $[[S]] \in CR_2 \iff\ [[forall a, (# a in S implies wn a #)]]$
  \end{itemize}
\end{definition}

@@

The automation provided by CoqHammer not only gives us a proof that is shorter
and more resiliant to changes, but also gives useful documentation for readers
who wish to understand the mechanized proof. Although automation performs
extensive search, it does not use lemmas or invert derivations that are not
specified in the \texttt{use} or \texttt{inv} flag.
